---
title: "Preliminary Analysis: GSE295191"
author: "ZB3101 CA2"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
    theme: cosmo
execute:
  warning: false
  message: false
---

# Import packages

```{r import-packs}
library(GEOquery)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(tidyr)
library(stringr)
# BiocManager::install('GEOquery')
```

# Check that both FASTQ samples are present

```{r check-files}
files <- list.files("data", pattern = "^(SRR33191495|SRR33191498).*\\.fastq$")

if (length(files) != 4) {
  stop("Required files are missing in data/")
} else {
  print(files)
  print("required files are present")
}
```

# Pull metadata from GEO

```{r geo-metadata}
# get metadata from GEO
gse   <- getGEO("GSE295191", GSEMatrix = TRUE, getGPL = FALSE)
pheno <- pData(phenoData(gse[[1]]))

# Clean up condition / replicate information for a readable table
meta_clean <- pheno |>
  mutate(
    Condition = case_when(
      str_detect(title, regex("dmso",      ignore_case = TRUE)) ~ "DMSO (Control)",
      str_detect(title, regex("ponatinib", ignore_case = TRUE)) ~ "Ponatinib 10 nM",
      TRUE ~ NA_character_
    ),
    Replicate = str_extract(title, "\\d+$")
  ) |>
  select(geo_accession, title, source_name_ch1, Condition, Replicate) |>
  rename(
    `GEO Accession` = geo_accession,
    `Sample Title`  = title,
    `Source`        = source_name_ch1
  )

# Display cleaned metadata
meta_clean |>
  kable(align = "llllc") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  column_spec(4, bold = TRUE,
              color = ifelse(meta_clean$Condition == "DMSO (Control)",
                             "#1a6faf", "#a63603"))
```

---

# Quality Control with FastQC

## 2.1 Sample selection

Two samples (one per condition) were selected for preliminary QC:

```{r sample-selection}
qc_samples <- tibble(
  `SRA Run`   = c("SRR33191495", "SRR33191498"),
  Condition   = c("DMSO (Control)", "Ponatinib 10 nM"),
  `Read 1`    = c("SRR33191495_1.fastq", "SRR33191498_1.fastq"),
  `Read 2`    = c("SRR33191495_2.fastq", "SRR33191498_2.fastq")
)

qc_samples |>
  kable(align = "llll") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## 2.2 Run FastQC

FastQC is run on both mates (R1 + R2) of each sample. Results are written to
`results/fastqc/` and the per-sample zip archives are unpacked so that the PNG
figures can be embedded below.

```{bash run-fastqc}
#| eval: false

mkdir -p results/fastqc
fastqc data/*.fastq* -o results/fastqc
```

```{r unzip-fastqc}
fastqc_dir <- "results/fastqc"

if (dir.exists(fastqc_dir)) {
  zips <- list.files(fastqc_dir, pattern = "_fastqc\\.zip$", full.names = TRUE)
  for (z in zips) {
    out_dir <- sub("\\.zip$", "", z)
    if (!dir.exists(out_dir)) unzip(z, exdir = fastqc_dir)
  }
}
```

## 2.3 FastQC summary across the two samples

```{r fastqc-summary-table}
summary_files <- list.files(
  "results/fastqc", pattern = "summary\\.txt$",
  recursive = TRUE, full.names = TRUE
)

if (length(summary_files) > 0) {
  summaries <- lapply(summary_files, function(f) {
    df <- read.delim(f, header = FALSE,
                     col.names = c("Status", "Module", "File"))
    df$Sample <- sub("_fastqc.*", "", basename(f))
    df
  }) |> bind_rows()

  summaries |>
    select(Sample, Module, Status) |>
    pivot_wider(names_from = Sample, values_from = Status) |>
    kable(align = "lll") |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE)
} else {
  cat("*FastQC summary will appear here after the FastQC bash chunk is run.*")
}
```

---

## 2.4 DMSO control — SRR33191495 (Read 1)

```{r fig-dmso-quality}
#| fig-cap: "Per-base sequence quality — DMSO (SRR33191495_1)."
#| out-width: "85%"
img <- "results/fastqc/SRR33191495_1_fastqc/Images/per_base_quality.png"
if (file.exists(img)) knitr::include_graphics(img) else
  cat("*Figure appears here after FastQC is executed.*")
```

```{r fig-dmso-dup}
#| fig-cap: "Sequence duplication levels — DMSO (SRR33191495_1)."
#| out-width: "85%"
img <- "results/fastqc/SRR33191495_1_fastqc/Images/duplication_levels.png"
if (file.exists(img)) knitr::include_graphics(img) else
  cat("*Figure appears here after FastQC is executed.*")
```

```{r fig-dmso-adapter}
#| fig-cap: "Adapter content — DMSO (SRR33191495_1)."
#| out-width: "85%"
img <- "results/fastqc/SRR33191495_1_fastqc/Images/adapter_content.png"
if (file.exists(img)) knitr::include_graphics(img) else
  cat("*Figure appears here after FastQC is executed.*")
```

---

## 2.5 Ponatinib — SRR33191498 (Read 1)

```{r fig-pon-quality}
#| fig-cap: "Per-base sequence quality — Ponatinib (SRR33191498_1)."
#| out-width: "85%"
img <- "results/fastqc/SRR33191498_1_fastqc/Images/per_base_quality.png"
if (file.exists(img)) knitr::include_graphics(img) else
  cat("*Figure appears here after FastQC is executed.*")
```

```{r fig-pon-dup}
#| fig-cap: "Sequence duplication levels — Ponatinib (SRR33191498_1)."
#| out-width: "85%"
img <- "results/fastqc/SRR33191498_1_fastqc/Images/duplication_levels.png"
if (file.exists(img)) knitr::include_graphics(img) else
  cat("*Figure appears here after FastQC is executed.*")
```

```{r fig-pon-adapter}
#| fig-cap: "Adapter content — Ponatinib (SRR33191498_1)."
#| out-width: "85%"
img <- "results/fastqc/SRR33191498_1_fastqc/Images/adapter_content.png"
if (file.exists(img)) knitr::include_graphics(img) else
  cat("*Figure appears here after FastQC is executed.*")
```

---

## 2.6 Interpretation

### Per-base sequence quality
Both samples are expected to show Phred scores ≥ 30 across most cycles (typical
for Illumina NovaSeq 6000 output). A modest drop at the 3′ end is normal. If
any cycles fall below Phred 20, quality trimming will be necessary before
alignment.

### Sequence duplication
Activated CD8+ T cells have a relatively low-complexity transcriptome (a
handful of cytokines, effector molecules and ribosomal genes dominate), so
duplication rates of ~20–40% are expected and biologically explainable.
Duplication > 60% would suggest PCR over-amplification and warrant further
investigation. We do **not** remove duplicates for bulk RNA-seq by default,
since duplicate reads arising from highly-expressed transcripts are real
signal; they are only inspected as a QC flag.

### Adapter content
Illumina TruSeq/Nextera adapters may appear at the 3′ end of short inserts.
NovaSeq libraries generally have low adapter contamination, but any detectable
adapter (especially > 5%) means we must include an adapter-trimming step.

### Consequences for the analysis pipeline

```{r qc-decision-table}
tribble(
  ~Observation,                              ~`Pipeline decision`,
  "Quality drop at 3′ end (common, mild)",   "fastp with --cut_tail --cut_tail_mean_quality 20, keep reads ≥ 50 bp",
  "Adapter content visible in either sample","fastp --detect_adapter_for_pe (auto-detect Illumina TruSeq)",
  "Duplication 20–40% (expected)",           "No pre-alignment deduplication; inspect duplicate rate post-STAR only",
  "Both samples pass all core modules",      "Proceed with all 6 samples to STAR alignment (GRCm39 + GENCODE M35)",
  "Either sample fails quality or GC modules","Flag for possible exclusion; re-check in PCA of raw counts"
) |>
  kable(align = "ll") |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE, width = "45%") |>
  column_spec(2, width = "55%")
```

**Summary:** Provided both samples pass the per-base quality, duplication and
adapter modules, the downstream plan is:

1. **Trimming:** `fastp` with default quality cutoff 20, min length 50 bp,
   paired-end adapter detection enabled.
2. **Alignment:** `STAR 2.7+` against GRCm39 with GENCODE M35 annotation,
   `--sjdbOverhang 149` (read length − 1, assuming 150 bp reads).
3. **Quantification:** `featureCounts` (Subread) with strand-specific mode
   confirmed from library prep kit (likely `-s 2` for TruSeq stranded).
4. **Differential expression:** `DESeq2` with design `~ Condition` on 3 vs. 3
   replicates. Shrinkage estimators (`lfcShrink(type = "apeglm")`) will be
   used given the low replicate count.

If FastQC flags one sample as an outlier (e.g. unusual GC curve or extreme
duplication), that sample will be carried forward but inspected in a PCA of
`vst`-normalised counts before committing to the final DE model.

---

# Session information

```{r session-info}
sessionInfo()
```
